### 2023-03-05 uvp6_part.qmd
remaster uvp6 article

# UVP6 embedded classification algorithm

@@

Millions of images of oceanic particles have been acquired with ship-based imaging instruments. The task of labeling images is time consuming but necessary to train classification models using artificial intelligence algorithms. Autonomous Underwater Vehicles (AUVs) such as Biogeochemical-Argo floats (BGC-Argo) have revolutionized our understanding of oceanic systems but they are flawed by their battery life and by the time they can spend at the ocean surface to transmit their data. In this paper, we show how we developed the first zooplankton classification model for AUVs using the Underwater Vision Profiler 6 (UVP6). To this end, we used a features-based classifier called eXtreme Gradient Boosting (XGBoost) adapted to specific energy constraints imposed by profiling floats and we compared its performance with a state-of-the-art convolutional neural network as if we were not energy-limited. In addition, we show how we dealt with a highly imbalanced data set towards marine snow (\> 80% of detritus) and what were the compromises to improve the classification scores of biological classes. Finally, we show the classification report (precision and recall for each label) of the final model and use it to analyse some preliminary results from BGC-Argo floats deployed in the Labrador Sea.

@@

Inside the ocean, a cohort of diverse organisms known as zooplankton thrive. Those tiny beasts are so important for our oceans [@Steinberg2017-ri] that we decided to take pictures of them [@Picheral2010-to; @Gorsky2010-ca; @Cowen2008-wq; @Benfield2007-lt; @Lombard2019-ld]. Though, the amount of images taken by those imaging instruments has become so huge that it is now a massive work to classify them by hand. Fortunately, powerful machine learning techniques have been implemented to tackle the issue of automatic plankton image classification [@Luo2018-oo].

@@

The discussion around automatic plankton classification is a key topic in oceanography and the development of convolutional neural networks (CNNs) within that frame has already lead to promising results [@Luo2018-oo; @Cui2018-nt; @Dai2017-ex; @Campbell2020-cq; @Schroder2019-um; @Bochinski2019-af; @Lee2016-jl; @Cheng2019-ar; @Guo2021-xi; @Ellen2019-nh; @Orenstein2017-ya]. Nevertheless, real-time image processing done at sea is far less mentioned to our knowledge. Usually, images are brought back to the lab and classified with CNNs which, in comparison with features-based classifiers, that is machine learning algorithms using a set of geometric features extracted from the image (e.g. area, eccentricity, grey levels) have outperformed the latter for plankton recognition [@Ellen2019-nh]. Though, it comes at a cost. CNNs are energy consuming hence their implementation for embedded applications can be limited by power consumption [@Biswas2018-uw; @Khabbazan2019-oi; @Oh2017-kj].

This is of particular importance for autonomous underwater vehicles (AUVs) and especially for the next generation of Biogeochemical-Argo (BGC-Argo) floats that will be equipped with the Underwater Vision Profiler 6 (UVP6), the portable version of the UVP5 [@Picheral2010-to] to better understand the carbon cycle [@Xing2018-zo]. The lifetime of floats is limited by their batteries hence by the power consumed by its sensors. Therefore, using an embedded CNN to classify plankton images or sending images through iridium transmission to be classified in the lab would not be possible for economical (cost of iridium transmission) and technical reasons. Typically, a BGC-Argo float samples the water column between the surface and 1000 m, it would therefore have stay at the surface for a time long enough to transmit all images collected after each profile, consuming energy proportionally to the size of the image set while being subject to biofouling or unpredictable hazards. For all these reasons, the development of an embedded plankton recognition algorithm on board the UVP6 is only possible with a features-based classifier at the moment.

In the literature, such algorithms (e.g. Decision Tree (DT), Random Forest (RF), Support Vector Machine (SVM)) have already been used for plankton recognition (see Table 1 of @Gonzalez2017-tr for an exhaustive list). For instance, @Blaschko2005-ep obtained an accuracy of 71% using a SVM on a set of 982 images among 13 classes using a 10-fold cross-validation (CV), @Sosik2007-cj obtained a global accuracy of 88% using a SVM on a test set of 3300 images covering 22 classes, @Bell2008-yo obtained 82% of accuracy using a 10-fold CV RF algorithm on 53 classes with 10 to 30 images per class. However, it should be noted that this accuracy dropped to 63% after the removal of non-biological classes such as detritus and fiber. For bigger training sets, which will be the case in this study, @Orenstein2017-ya\]trained a RF model on approximately 650.000 images across 70 classes for an accuracy of 91% on the test set (2.75 million images). Though, one class in that test set contained 1.6 million images and if the classifier had put every object in that class, the accuracy would have been around 60% so caution must be applied when interpreting performance metrics. @Faillettaz2016-wr trained a RF model on 6000 images across 14 classes, reaching an accuracy of 56% on a test set of 1.5 million objects, discarding those below a certain probability threshold. This allowed to improve the global accuracy by 16% but resulted in the loss of 72% of objects in the predicted data set.

In fine, it is difficult to compare all methods due to the diversity of imaging instruments hence the variety of image formats (e.g. resolution), the number of classes (tradeoff between the model performance and the number of classes as illustrated in @Fernandes2009-hw, the training set (e.g. number of images per class, diversity, balance), the scientific objective (e.g. precision-recall tradeoff), the type and number of features and the algorithm used. In many cases, RF seemed to be the best classifier [@Gorsky2010-ca; @Faillettaz2016-wr] but in embedded applications, energy constraints need to be taken into account. Therefore, the model needs to be small and the classification should be easy and fast. Gradient Tree Boosting (GTB) seems to offer this compromise when compared to RF (model size) and SVM (computation complexity). For these reasons, we chose XGBoost [@Chen2016-xt], an optimized implementation of GTB, also known to be one of the best assets to win Kaggle's machine learning competitions.

In this paper, we present the results of an embedded zooplankton recognition algorithm using XGBoost on UVP5 data. We show how we built the embedded recognition (ER) model under strict technical constraints, what were our compromises and what techniques we used to improve the classification. We also compare the performances of our features-based algorithm with a state-of-the-art neural network as if we were not limited by energy constraints. Finally, we apply the method and follow the recommendations learnt from the preliminary UVP5 model to the UVP6 data set to build an ER model that will be included in the UVP6 on future BGC-Argo floats.

@@

The UVP6 is the miniaturized version of the UVP5 [@Picheral2021-sx]. It is smaller, it uses less energy and it has been specifically developed to be mounted on AUVs such as BGC-Argo floats with the aim of providing data of similar quality than from the UVP5. There are 2 versions: the LP (low power) and the HF (high frequency). The former uses less energy (lower image acquisition rate) and will be deployed on AUVs while the latter is designed to be used on CTD rosettes and cruising AUVs [@Picheral2021-sx].

@@

The full UVP5 SD data set comprises 3.578.901 validated images in 2273 profiles. The full UVP6 LP data set comprises 634.459 validated images in XX profiles (all UVP6 profiles were not entirely validated). Each image was segmented to find the main object (i.e. the biggest connected region in pixels). From it, we extracted 55 features describing this object in terms of shape (e.g. pixel area, eccentricity), grey level and mostly statistical moments (all features are available in **Fig. SXX**). For technical reasons, all UVP6 images with a pixel area below 79 pixels were removed (7% of images).

Validated objects were manually classified into specific taxa belonging to well known biological groups such as Rhizaria, Copepoda or Cnidaria to name a few. However, the maximum number of classes (also referred to as labels or taxa) for the ER was fixed to 40 in order to limit the size of the ER model and the volume of data that needs to be transmitted. We chose to aggregate classes iteratively to find the best number of classes for the ER. The first level (ie. all labels (\> 40) are retained) served as a baseline for a prior model, without making any hypotheses. The next level was used as a first aggregation step based on the number of objects in each class and the purity of the class. Below a given threshold, rare classes were merged with their most look-alike taxa and parents on the phylogenetic tree were removed when their children were abundant (i.e. parents could not be classified to a deeper taxonomic level, likely causing confusion with their children during training). The second level of aggregation was based on statistical metrics post training (see hereafter how a model is trained), meaning that we looked at the confusion matrix and the classification report (e.g. **Fig. XX and XX**, respectively) to see where classes did not perform well such that they had to be merged to improve the classification score. This last step was repeated until an agreement was found. Thus, classes were merged only if they were confused amongst the same biological group on the phylogenetic tree or if they had a similar ecological role. This involves that some classes end up with a variety of organisms that do not necessarily show similar morphological traits (e.g. other\<living). It is definitely not ideal but we were bound to keep as many biological classes as possible to counterbalance non-living classes.

In machine learning tasks, it is common to split the data into a training set, a validation set and a test set. The validation set is usually used to assess the quality of the model by tuning hyperparameters while avoiding overfitting (stopping the training before the model is too acquainted with the data). Ideally, the test set is not to be used during any step except at the very end where it is used to assess the global quality of the final model. In our case, both datasets (UVP5 and UVP6) are very imbalanced towards few non-biological classes (mostly detritus). As a result, in order for the model to see as much biological data as possible during training, we did not use a validation set but instead we did a threefold cross-validation.

One objective of the ER on board the UVP6 is to be used on AUVs such as BGC-Argo floats to study the vertical migration pump [@Claustre2021-tx]. Hence, we built the test set with entire UVP5 profiles going at least to 1000 m depth, the typical parking depth of BGC-Argo floats. Those profiles were also chosen to be representative of the global ocean (**see Fig. XX**). For UVP6 data, we did not select entire profiles because we were limited by the number of images in comparison with the UVP5 to build the biggest training set possible (80% of detritus in the UVP6 data set). Therefore, we splitted the data set in a 10-90 % stratified test-train fashion regardless of specific profiles.

Following the previous steps, the UVP5 test and training sets are composed, respectively, of 315.277 and 3.023.506 images while the UVP6 has 59.054 and 531.483 images for its test and training sets.

@@

GTB generates a sequence of trees, each one trying to cope with the errors made by their predecessors. XGBoost is designed to be an optimized and regularized version of GTB. To learn from the data and derive a model, XBoost first minimizes a regularized objective function $L$

@@

where $l$ is a loss function that measures the difference between the prediction $\hat{y}_{i}$ and the true value $y_{i}$ and $\Omega$ is the regularization term, for each $k$ decision tree $f_{k}$. $\Omega$ is then splitted into two terms where $\gamma T$ penalizes the complexity of a tree proportionally to the number of leaves ($T$) and where the L2 norm penalizes the weights ($w$) of the leaves to avoid over-fitting (both $\gamma$ and $\lambda$ are hyperparameters that can be tuned by the user). By doing so, XGBoost encourages both a simple and predictive model. From @eq-objective-function-xgboost, a scoring function that measures the quality of a tree structure can be derived (see Eq. 6 in @Chen2016-xt for details). However, in practice, it is not possible to test all suitable trees because there are infinite possibilities. Instead, the algorithm starts from a single leaf and iteratively adds branches to the tree, searching for the best split options.

To build our ER model, we first did a threefold cross-validation with XGBoost to find the optimal number of boosting rounds (i.e. $k$) before the model overfits. For this, we looked at the minimum of the multi-class log loss curve (**Fig. S3**), then we trained the ER model with the entire training set at this optimal number with specific hyperparameters (ie. learning rate ($\eta$), maximum depth of a tree ($D_{max}$) and weight sensitivity ($\omega$). The optimal couple ($\eta$, $D_{max}$) was determined using a customary grid search (**see Fig. 3**) while $\omega$ was chosen empirically (see hereafter).

To cope with our unbalanced dataset (more than 70% of detritus in the UVP5 dataset), several options were available. We could artificially increase the number of biological data with the well known SMOTE oversampling technique [@Chawla2002-kg; @Fernandez2018-ft], decrease the number of detritus (random undersampling, keep only detritus close to unsupervised cluster centroids, etc.) or add a weight to each label during training to nudge the model towards under-represented classes (i.e. biological classes). In order to compare a features-based classifier with a state-of-the-art CNN, we chose the latter.

>>>>

# UVP6 embedded classification algorithm {#sec-uvp6}

@@

Millions of images of oceanic particles have been acquired with ship-based imaging instruments. The task of labeling images is time consuming but necessary to train classification models using artificial intelligence algorithms. Autonomous Underwater Vehicles (AUVs) such as Biogeochemical-Argo floats (BGC-Argo) have revolutionized our understanding of oceanic systems but they are flawed by their battery life and by the time they can spend at the ocean surface to transmit their data. In this paper, we show how we developed the first zooplankton classification model for AUVs using the Underwater Vision Profiler 6 (UVP6). To this end, we used a features-based classifier called eXtreme Gradient Boosting (XGBoost) adapted to specific energy constraints imposed by profiling floats and we compared its performance with a state-of-the-art convolutional neural network as if energy was not a limiting factor. In addition, we show how we dealt with a highly imbalanced data set towards marine snow (\> 80% of detritus) and what were the compromises to improve the classification scores of biological classes. Finally, we show the classification report (precision and recall for each label) of the final classification model and use it to analyse some preliminary results from BGC-Argo floats deployed in the Labrador Sea.

@@

Inside the ocean, a cohort of diverse organisms known as zooplankton thrive. Those tiny beasts are so important for our oceans [@Steinberg2017-ri] that we decided to take pictures of them [@Picheral2010-to; @Gorsky2010-ca; @Cowen2008-wq; @Benfield2007-lt; @Lombard2019-ld]. Nonetheless, the amount of images taken by those imaging instruments has become so huge that it is now a massive work to classify them by hand. Fortunately, powerful machine learning techniques have been implemented to tackle the issue of automatic plankton image classification [@Luo2018-oo].

@@

The discussion around automatic plankton classification is a key topic in oceanography and the development of convolutional neural networks (CNNs) within that frame has already lead to promising results [@Luo2018-oo; @Cui2018-nt; @Dai2017-ex; @Campbell2020-cq; @Schroder2019-um; @Bochinski2019-af; @Lee2016-jl; @Cheng2019-ar; @Guo2021-xi; @Ellen2019-nh; @Orenstein2017-ya]. Nevertheless, real-time image processing done at sea is far less mentioned to our knowledge. Usually, images are brought back to the lab and classified with CNNs which, in comparison with features-based classifiers, that is machine learning algorithms using a set of features extracted from the image (e.g. area, eccentricity, grey levels) have outperformed the latter for plankton recognition [@Ellen2019-nh]. Though, it comes at a cost. CNNs are energy consuming hence their implementation for embedded applications can be limited by power consumption [@Biswas2018-uw; @Khabbazan2019-oi; @Oh2017-kj].

This is of particular importance for autonomous underwater vehicles (AUVs) and especially for a new version of Biogeochemical-Argo (BGC-Argo) floats that will be equipped with the Underwater Vision Profiler 6 (UVP6), the portable version of the UVP5 [@Picheral2010-to] to better understand the carbon cycle [@Xing2018-zo]. The lifetime of floats is limited by their batteries hence by the power consumed by its sensors. Therefore, using an embedded CNN to classify plankton images or sending images through iridium transmission to be classified in the lab would not be possible for economical (cost of iridium transmission) and technical reasons. Typically, a BGC-Argo float samples the water column between the surface and 1000 m, it would therefore have to stay at the surface for a time long enough to transmit all images collected after each profile, consuming energy proportionally to the size of the image set while being subject to biofouling or unpredictable hazards. For all these reasons, it was decided to use a features-based classifier for the development of the embedded plankton recognition algorithm on board the UVP6.

In the literature, such algorithms (e.g. Decision Tree (DT), Random Forest (RF), Support Vector Machine (SVM)) have already been used for plankton recognition (see Table 1 of @Gonzalez2017-tr for an exhaustive list). For instance, @Blaschko2005-ep obtained an accuracy of 71% using a SVM on a set of 982 images among 13 classes using a 10-fold cross-validation (CV), @Sosik2007-cj obtained a global accuracy of 88% using a SVM on a test set of 3300 images covering 22 classes, @Bell2008-yo obtained 82% of accuracy using a 10-fold CV RF algorithm on 53 classes with 10 to 30 images per class. However, it should be noted that this accuracy dropped to 63% after the removal of non-biological classes such as detritus and fiber. For bigger training sets, which will be the case in this study, @Orenstein2017-ya trained a RF model on approximately 650.000 images across 70 classes for an accuracy of 91% on the test set (2.75 million images). Though, one class in that test set contained 1.6 million images and if the classifier had put every object in that class, the accuracy would have been around 60% so caution must be applied when interpreting performance metrics. @Faillettaz2016-wr trained a RF model on 6000 images across 14 classes, reaching an accuracy of 56% on a test set of 1.5 million objects, discarding those below a certain probability threshold. This allowed to improve the global accuracy by 16% but resulted in the loss of 72% of objects in the predicted data set.

It is quite difficult to compare all methods due to the diversity of imaging instruments hence the variety of image formats (e.g. resolution), the number of classes (trade off between the model performance and the number of classes as illustrated in @Fernandes2009-hw, the training set (e.g. number of images per class, diversity, balance), the scientific objective (e.g. precision-recall trade off), the type and number of features and the algorithm used. In many cases, RF seemed to be the best classifier [@Gorsky2010-ca; @Faillettaz2016-wr] but in embedded applications, energy constraints need to be taken into account. Therefore, the model needs to be small and the classification should be easy and fast. Gradient Tree Boosting (GTB) seems to offer this compromise when compared to RF (model size) and SVM (computation complexity). For these reasons, we chose XGBoost [@Chen2016-xt], an optimized implementation of GTB, also known to be one of the best assets to win Kaggle's machine learning competitions.

In this paper, we present the results of an embedded zooplankton recognition algorithm using XGBoost on UVP5 data. We show how we built the embedded recognition (ER) model under strict technical constraints, what were our compromises and what techniques we used to improve the classification. We also compare the performances of our features-based algorithm with a state-of-the-art neural network as if energy was not a limiting factor. Finally, we apply the method and follow the recommendations learnt from the preliminary UVP5 model to the UVP6 data set to build the ER model that will be included in the UVP6 for AUVs.

@@

The UVP6 is the miniaturized version of the UVP5 [@Picheral2021-sx]. It is smaller, it uses less energy and it has been specifically developed to be mounted on AUVs such as BGC-Argo floats with the aim of providing data of similar quality than from the UVP5 (@fig-uvp-pipeline). There are 2 versions: the LP (low power) and the HF (high frequency). The former uses less energy (lower image acquisition rate) and will be deployed on AUVs while the latter is designed to be used on CTD rosettes and cruising AUVs [@Picheral2021-sx].

![Scheme of the UVP6 pipeline with the processing done in the lab to create the ER model and the processing time (in ms) of the main steps when the UVP6 is deployed on board an AUV. kpx means kilo pixel.](image/uvp6/fig_pipeline_UVP6.jpg){#fig-uvp-pipeline}

@@

The full UVP5 SD data set is composed of 3.578.901 validated images whereas the full UVP6 LP data set has 634.459 validated images. Each image was segmented to find the main object (i.e. the biggest connected region in pixels). From it, 55 features were extracted to describe this object in terms of shape (e.g. pixel area, eccentricity), gray level and mostly statistical moments (see @tbl-features). For technical reasons, all UVP6 images with a pixel area below 79 pixels were removed (7% of images).

```{r, features, dpi = 300}
#| label: tbl-features
#| tbl-cap: List of 55 features (abbreviations used in our python package UVPec (https://github.com/ecotaxa/uvpec) and their descriptions) extracted from each object based on the object geometrical characteristics, the pixel gray values histogram and statistical moments.

# library(gt)
# library(gtExtras)
#library(flextable)
library(kableExtra)
library(tibble)

# object_attributes <- tibble(attr = c('area', 'width', 'height', 'x','y','xm','ym','major','minor', 'angle', 'eccentricity', 'esd', 'bbox_area', 'extent'), comment = c('number of pixels', 'bounding box width', 'bounding box height', 'local centroid (x direction)', 'local centroid (y direction)', 'weighted local centroid (x direction)', 'weighted local centroid (y direction)', 'major axis', 'minor axis', 'orientation angle','-', 'equivalent spherical diameter', 'bounding box area', 'pixel ratio between object and bounding box'))
# 
# hist_attributes <- tibble(attr = c('mean', 'stddev', 'mode', 'min', 'max', 'intden', 'median','histcum1', 'histcum3', 'vrange', 'meanpos', 'cv', 'sr'), comment = c('-', 'standard deviation', '-', '-', '-', 'sum of all pixel values', '-', 'first quartile', 'third quartile', 'vmax - vmin', '(vmax - vmin)/range', '100*(stddev/mean)', '100*(stddev/vrange)'))
# 
# stat_moment <- tibble(attr = c('hu_moment-i (i: 1 to 7)', 'gray_hu_moment-i (i: 1 to 7)',
#                                'central_moment-2-0',
#                                'central_moment-1-1',
#                                'central_moment-0-2',
#                                'central_moment-3-0',
#                                'central_moment-2-1',
#                                'central_moment-1-2',
#                                'central_moment-0-3',
#                                'gray_central_moment-2-0',
#                                'gray_central_moment-1-1',
#                                'gray_central_moment-0-2',
#                                'gray_central_moment-3-0',
#                                'gray_central_moment-2-1',
#                                'gray_central_moment-1-2',
#                                'gray_central_moment-0-3'),
#                       comment = c('Hu moments (translation, scale and rotation invariant)', 'weighted Hu moments', 'central moments (translation invariant)', '-', '-', '-', '-', '-', '-', 'weighted central moments', '-', '-', '-', '-', '-', '-'))

object_attributes <- tibble(attr = c('area', 'width', 'height','major','minor', 'angle', 'eccentricity', 'esd', 'bbox_area', 'extent'), comment = c('number of pixels', 'bounding box width', 'bounding box height', 'major axis', 'minor axis', 'orientation angle','eccentricity', 'equivalent spherical diameter', 'bounding box area', 'pixel ratio between object and bounding box'))

hist_attributes <- tibble(attr = c('mean', 'stddev', 'mode', 'min', 'max', 'x','y','xm','ym','intden', 'median','histcum1', 'histcum3', 'vrange', 'meanpos', 'cv', 'sr'), comment = c('mean pixel intensity', 'standard deviation of pixel intensity', 'pixel intensity mode', 'min pixel intensity', 'max pixel intensity', 'local centroid (x direction)', 'local centroid (y direction)', 'weighted local centroid (x direction)', 'weighted local centroid (y direction)', 'sum of all pixel values', 'median pixel intensity', 'pixel intensity first quartile', 'pixel intensity third quartile', 'vmax - vmin', '(vmax - vmin)/range', '100*(stddev/mean)', '100*(stddev/vrange)'))

stat_moment <- tibble(attr = c('hu_moment-i (i: 1 to 7)', 'gray_hu_moment-i (i: 1 to 7)',
                               'central_moment-i-j (2-0;1-1;0-2;3-0;2-1;1-2;0-3)',
                               'gray_central_moment-i-j (2-0;1-1;0-2;3-0;2-1;1-2;0-3)'),
                      comment = c('Hu moment', 'weighted Hu moment', 'central moment', 'weighted central moment'))

all_features <- rbind(object_attributes, hist_attributes, stat_moment)
colnames(all_features) <- c('Features name', 'Description')

# gt(all_features) |> gt_highlight_rows(rows = 1:13,
#                                       fill = 'lightgrey')
#all_features |> flextable() |> color(i = 1:13, color = 'lightgrey')
# kable(head(iris, 5), align = 'c', booktabs = TRUE) %>%
#   row_spec(1, bold = TRUE, italic = TRUE) %>% 
#   row_spec(2:3, color = 'white', background = 'black') %>%
#   row_spec(4, underline = TRUE, monospace = TRUE) %>% 
#   row_spec(5, angle = 45) %>% 
#   column_spec(5, strikeout = TRUE)

# kable(all_features) |> row_spec(1:14, background = '#F2F2F2') |> 
#   row_spec(28:43, background = "#F2F2F2") |> row_spec(0, bold=T, align = 'c')

kable(all_features) |> row_spec(0, bold=T, align = 'c')

```

Validated objects were manually classified into specific taxa belonging to well known biological groups such as Rhizaria, Copepoda or Cnidaria to name a few. However, the maximum number of classes (also referred to as labels or taxa) for the ER was fixed to 40 in order to limit the size of the ER model and the volume of data that needs to be transmitted. We chose to aggregate classes iteratively to find the best number of classes for the ER. The first level (i.e. all labels (\> 40) are retained) served as a baseline for a prior model, without making any hypotheses. The next level was used as a first aggregation step based on the number of objects in each class and the purity of the class. Below a given threshold, rare classes were merged with their most look-alike taxa and parents on the phylogenetic tree were removed when their children were abundant (i.e. parents could not be classified to a deeper taxonomic level, likely causing confusion with their children during training). The second level of aggregation was based on statistical metrics post training (see hereafter how a model is trained), meaning that we looked at the confusion matrix and the classification report (@fig-confusion-matrix-uvp5 and @tbl-classification-report-uvp5-xgboost-cnn, respectively) to find which classes did not perform well such that they had to be merged to improve the classification score. This last step was repeated until an agreement was found. Thus, classes were merged only if they were confused amongst the same biological group on the phylogenetic tree or if they had a similar ecological role. This involves that some classes end up with a variety of organisms that do not necessarily show similar morphological traits (e.g. other\<living). It is definitely not ideal but we were bound to keep as many biological classes as possible to counterbalance non-living classes.

<!-- ![Confusion matrix (normalized on true labels) of the best UVP5 XGBoost model.](image/uvp6/confusion_matrix_best_model_uvp5.jpg){#fig-confusion-matrix-uvp5-xgboost} -->

```{r, confusion matrix CNN UVP5, fig.height = 6}
#| label: fig-confusion-matrix-uvp5
#| fig-cap: Confusion matrix (normalized on true labels) for the best UVP5 model trained with XGBoost and a CNN.

library(viridis)
library(gridExtra)

# define x and y axes
x <- c('Acantharia', 'Aulacanthidae', 'Aulosphaeridae', 'Copepoda', 'Trichodesmium', 
            'artefact', 'bubble', 'darksphere', 'detritus', 'fiber', 'other<living')

y <- factor(x, levels = rev(x))

# create grid
data <- expand.grid(X=x, Y=y, stringsAsFactors = T)

# gray levels
gray_cnn_uvp5 <- c(38,252,254,254,242,254,255,255,251,253,250,
           249,61,252,255,241,253,255,255,247,255,251,
           251,250,17,255,255,253,255,255,250,255,252,
           255,255,255,61,255,255,255,255,214,253,252,
           254,255,255,255,0,255,255,255,245,253,255,
           255,255,255,255,255,92,255,255,187,254,254,
           255,255,255,255,253,253,0,255,246,255,255,
           251,248,253,255,242,254,252,109,222,255,252,
           255,255,255,253,251,248,255,255,17,251,254,
           255,255,255,255,248,253,255,255,189,108,254,
           250,252,255,247,253,247,255,255,230,249,99)

# reverse grayscale so that 0 becomes 255 and 255 becomes 0 and normalize to 1
gray_cnn_uvp5 <- abs(gray_cnn_uvp5 - 255)/255

# add gray values in the grid
data$gray_level <- gray_cnn_uvp5

# add group for facet plot
data <- data |> mutate(group = 'CNN')

right <- ggplot(data, aes(X, Y, fill= gray_level)) + geom_tile() + scale_fill_viridis(option = "F", direction = -1) +
  labs(x = "Predicted labels", y = "True labels") + theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# do the same for the xgboost confusion matrix
gray_xgboost_uvp5 <- c(66,252,249,255,250,255,255,255,233,249,250,
                        253,62,248,255,248,244,255,255,244,255,249,
                        254,252,4,255,255,255,255,255,243,255,251,
                        255,255,255,126,255,253,255,255,147,253,252,
                        254,255,255,255,35,255,255,253,226,248,255,
                        255,255,255,255,255,35,255,255,213,255,255,
                        255,255,255,255,253,250,17,255,235,255,255,
                        254,253,253,255,251,254,255,65,215,255,252,
                        255,255,255,253,253,246,255,255,0,250,253,
                        255,255,255,254,250,252,255,255,154,120,255,
                        250,252,253,246,252,241,255,253,221,245,135)

gray_xgboost_uvp5 <- abs(gray_xgboost_uvp5 - 255)/255
data2 <- expand.grid(X=x, Y=y, stringsAsFactors = T)
data2 <- data2 |> mutate(gray_level = gray_xgboost_uvp5, group = 'XGBoost')

left <- ggplot(data2, aes(X, Y, fill= gray_level)) + geom_tile() + scale_fill_viridis(option = "F", direction = -1) +
  labs(x = "Predicted labels", y = "True labels") + theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# left / right

# plot facet
data <- rbind(data, data2)
data <- data |> mutate(group = factor(group, levels = c('XGBoost', 'CNN'))) 
ggplot(data, aes(X, Y, fill= gray_level)) + geom_tile() + scale_fill_viridis(option = "F", direction = -1) +
  labs(x = "Predicted labels", y = "True labels", fill = '') + theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + facet_wrap(~group, nrow = 2)

```

```{r, classification report XGBoost UVP5}
#| label: tbl-classification-report-uvp5-xgboost-cnn
#| tbl-cap: Precision, recall and f1-score for each class and globally (lightgray background) in the UVP5 model (XGBoost and CNN). avg average. avg stands for average.

labels <- c('Acantharea', 'Aulacanthidae', 'Aulosphaeridae', 'Copepoda', 'Trichodesmium', 
            'artefact', 'bubble', 'darsphere', 'detritus', 'fiber', 'other<living', 'accuracy',
            'macro avg', 'weighted avg', 'living macro avg', 'living weighted avg')

precision <- c(0.68,0.63,0.80,0.54,0.68,0.6,0.77,0.57,0.90,0.50,0.36,0.81,0.64,0.82,0.61,0.60)
recall <- c(0.68,0.69,0.84,0.50,0.75,0.75,0.81,0.68,0.86,0.52,0.48,0.81,0.69,0.81,0.66,0.65)
f1_score <- c(0.68,0.66,0.82,0.52,0.71,0.67,0.79,0.62,0.88,0.51,0.41,0.81,0.66,0.82,0.63,0.62)

precision_cnn <- c(0.60,0.61,0.89,0.56,0.56,0.63,0.76,.59,0.91,0.56,0.52,0.82,0.65,0.83,0.62,0.59)
recall_cnn <- c(0.79,0.73,0.87,0.72,0.89,0.63,0.89,0.58,0.87,0.58,0.61,0.82,0.74,0.82,0.74,0.78)
f1_score_cnn <- c(0.68,0.66,0.88,0.63,0.69,0.63,0.82,0.58,0.89,0.57,0.56,0.82,0.69,0.82,0.67,0.67)

cf_uvp5_xgboost_cnn <- tibble(labels, precision, recall, f1_score, precision_cnn, recall_cnn, f1_score_cnn)
colnames(cf_uvp5_xgboost_cnn) <- c('Labels', 'Precision', 'Recall', 'F1-score', 'Precision', 'Recall', 'F1-score')

kable(cf_uvp5_xgboost_cnn) |> row_spec(0, bold=T, align = 'c') |> #row_spec(1:16, align = 'c') |>
  row_spec(row = 12:16, background = '#F2F2F2') |> add_header_above(c(" " = 1, "XGBoost" = 3, "CNN" = 3))
  #column_spec(2, background = spec_color(cf_uvp5_xgboost$precision))
  #row_spec(12:16, background = '#F2F2F2')
```

In machine learning tasks, it is common to split the data into a training set, a validation set and a test set. The validation set is usually used to assess the quality of a model by tuning hyperparameters while avoiding overfitting (stopping the training before the model is too acquainted with the data). Ideally, the test set is not to be used during any step except at the very end where it is used to assess the global quality of the final model. In our case, both data sets (UVP5 and UVP6) are very imbalanced towards few non-biological classes (mostly detritus). As a result, in order for the model to see as much biological data as possible during training, we did not use a validation set but instead we did a threefold cross-validation.

One objective of the ER on board the UVP6 is to be used on AUVs such as BGC-Argo floats to study the vertical migration pump [@Claustre2021-tx]. Hence, we built the test set with entire UVP5 profiles going at least to 1000 m depth, the typical parking depth of BGC-Argo floats. Those profiles were also chosen to be representative of the global ocean (@fig-map-of-profiles). For UVP6 data, we did not select entire profiles because we were limited by the number of images in comparison with the UVP5 to build the biggest training set possible (80% of detritus in the UVP6 data set). Therefore, we splitted the data set in a 10-90 % stratified test-train fashion regardless of specific profiles.

Following the previous steps, the UVP5 test and training sets are composed of, respectively, 315.277 and 3.023.506 images while the UVP6 has 59.054 and 531.483 images for its test and training sets.

@@

GTB generates a sequence of trees, each one trying to cope with the errors made by their predecessors. XGBoost is designed to be an optimized and regularized version of GTB. To learn from the data and derive a model, XGBoost first minimizes a regularized objective function $L$

@@

where $l$ is a loss function that measures the difference between the prediction $\hat{y}_{i}$ and the true value $y_{i}$ and $\Omega$ is the regularization term, for each $k$ decision tree $f_{k}$. $\Omega$ is then splitted into two terms where $\gamma T$ penalizes the complexity of a tree proportionally to the number of leaves ($T$) and where the L2 norm penalizes the weights ($w$) of the leaves to avoid overfitting (both $\gamma$ and $\lambda$ are hyperparameters that can be tuned by the user). By doing so, XGBoost encourages both a simple and predictive model. From @eq-objective-function-xgboost, a scoring function that measures the quality of a tree structure can be derived (see Eq. 6 in @Chen2016-xt for details). However, in practice, it is not possible to test all suitable trees because there are infinite possibilities. Instead, the algorithm starts from a single leaf and iteratively adds branches to the tree, searching for the best split options.

To build the ER model, the first step was a threefold cross-validation using XGBoost parameterized with a set of given hyperparameters (i.e. learning rate ($\eta$), maximum depth of a tree ($D_{max}$) and weight sensitivity ($\omega$)). The second step consisted in the search of the optimal number of boosting rounds (i.e. $k$) before the model overfits. For this, we searched for the minimum value of the multi class log loss(also known as logistic loss or cross-entropy loss) evaluated at each boosting round. Using that optimal $k$ value, we then trained the ER model with the entire training set. Therefore, finding the best model consisted in finding the couple ($\eta$, $D_{max}$) that respected the energy constraints and outputed the best global metrics. The latters were determined using a customary grid search (@fig-grid-search) while $\omega$ was chosen empirically (see hereafter).

```{r, figure for grid search, fig.width=10, fig.height=6}
#| label: fig-grid-search
#| fig-cap: Grid search results with several classes of detritus (1, 3 or 5 classes). D$_{max}$ is depicted by the colour of the border of each shape (red or black).

library(tidyverse)
library(ggplot2)
library(stringr)
library(gghighlight)

# read all data
d <- bind_rows(
  read_csv('DATA/UVP6/results_grid_search_UVP5.csv') %>%
    filter(str_detect(folder_name, "_1000")) %>%
    mutate(learning_rate = learning_rate/10) %>%
    filter(learning_rate < 0.5, max_depth < 7, n_detritus < 6)
) %>%
  # remove metrics with less meaning
  #select(-ends_with("bio"), -mylogloss) %>%
  # convert to tall format
  gather(key="metric", val="val", accuracy:living_recall_w) %>%
  # reformat hyperparameters
  rename(n_det=n_detritus) %>%
  mutate(
    eta=factor(learning_rate),
    Dmax=factor(max_depth),
    n_det=factor(n_det),
  )

#https://stackoverflow.com/questions/64271350/changing-column-names-in-a-scatterplot-with-facet-wrap?noredirect=1&lq=1
panel_names <- c(`accuracy`="global accuracy",
                    `living_precision_w`="weighted living precision",
                    `living_recall_w`="weighted living recall",
                    `precision_macro`="macro precision",
                    `recall_macro`="macro recall")
